\mytitle

### 2015-07-24 MSc_diss.tex
Additions to desctiption of the chemical mechanism

\item The formation of the nitric acid ($HNO_3$) is a major loss process for $NO_x$ during daytime in the atmosphere (e.g. \citep{Holloway2010}). Being a 'sticky' molecule $HNO_3$ readily adsorbs to surfaces, particularly if there is water on the surface, and (deposits either in precipitation (wet deposition) or in dry form (dry deposition))/(undergoes dry and wet deposition). Because of the high deposition velocity, dry deposition of $HNO_3$ can be responsible for much of the removal of inorganic nitrogen from the troposphere \citep{Finlayson-Pitts2000}, which means that exclusion of $HNO_3$ (formation and removal)/(chemistry) from the chemical mechanism (potentially) leads to overproduction of ozone in the model, especially at high $NO_2$ concentrations because the competition between $NO_2$ photolysis and its reaction with $OH$ radical ($NO_2 + OH \xrightarrow{M} HNO_3$) is ignored.

@@

\item $NO_3$ radical rapidly photolyzes during the daylight hours, but in nighttime its concentration can increase to measurable levels \citep{Atkinson2008}. This increase in concentration and a lesser availability of the OH radical (which sources are mostly photolytic in nature) makes $NO_3$ radical a major contributor to the night chemistry of organics in the troposphere \citep{Finlayson-Pitts2000}. However, exclusion of $NO_3$ radical reactions from our chemical mechanism should not have a huge impact on the results anyway since the mechanism considers mainly decomposition of alkanes with which $NO_3$ radical reacts to a generally minor extent \citep{Atkinson2008}???.

@@

\item Peroxyacetyl nitrates ($RC(O)OONO_2$, PAN)
is important as a means of transporting NOx over large distances. It is quite stable at low temperatures but decomposes at higher temperatures to release NO2. Hence whether photolysis can compete with the thermal decomposition is of great interest.

The rate constant for PAN decomposition is strongly temperature dependent. As a result, this is important in the atmosphere because PAN acts as a reservoir for NO2. At low temperatures, much of the total NOx may be tied up in PAN; as the atmosphere warms, due to either duirnal temperature variations or the air mass being transported into warmer regions. the rate of PAN decomposition increases. This releases NO2, which can then form other secondary pollutants such as O3 and HNO3.

PAN's thermal decomposition releases both NO2 and an organic free radical (acetylperoxy radical), so that it can act as an NOx reservoir and ultimately as a source of OH in the dark.
The oxidation of organics produces a variety of free radicals, some of which can form peroxyacyl nitrates. For example, acetaldehyde is a classic precursor to the PAN through the following reaction sequence:
CH3CHO + OH = CH3CO + H2O; 
CH3CO + O2 + M = CH3C(O)OO + M;
CH3C(O)OO + NO2 + M = CH3C(O)OONO2 + M.
The key feature of the last reaction is that it is reversible.
Of all the possible fates for PANs in the atmosphere, thermal decomposition is usually the most important.
The strong dependence has important implications for the role of PAN. When it is formed at lower temperatures or is transported into colder regions, it is stabilized and acts as an NOx reservoir. When an air mass containing PAN is transported into warmer regions, however, the PAN decomposes, releasing both the NO2and the acetylperoxy free radical. If sufficient NO is present, the organic free radical reacts int the following manner:
CH3C(O)OO + NO = CH3C(O)O + NO2;
CH3C(O)O = CH3 + CO2
followed by the reaction of CH3 with O2, etc. to form HCHO and HO2, or HCHO, CH3OH and CH3OOH under low-NOx conditions. As a result, PAN should act as an accelerator for photochemical smog formation.
At lower temperatures such as those found in the Arctic, the majority  of NOy is often present in the form of PAN because of its thermal stability.
\end{itemize}

@@

! oxidation of methane ($CH_4$) and other alkanes, namely ethane ($C_2H_6$), propane ($C_3H_8$), n-butane ($nC_4H_{10}$), i-butane ($iC_4H_{10}$), n-pentane ($nC_5H_{12}$) and i-pentane ($nC_5H_{12}$)

@@

! Apart from methane in the model there are other nonmethane 
It is worth mentioning that only alkanes  
Due to purely theoretical purpose of this investigation the only sink for NOx considered in the mechanism is their loss due to formation of alkyl nitrates. It means that reactions involving formation of nitric acid ($HNO_3$) and peroxyacetyl nitrates (PAN) are not considered here, which could lead to discrepancies between our modelling results and those from other research studies.

>>>>

\item The formation of the nitric acid ($HNO_3$) is a major loss process for $NO_x$ during daytime in the atmosphere. Being a 'sticky' molecule $HNO_3$ readily adsorbs to surfaces, particularly if there is water on the surface, and deposits either in precipitation (wet deposition) or in dry form (dry deposition). Because of the high deposition velocity, dry deposition of $HNO_3$ can be responsible for much of the removal of inorganic nitrogen from the troposphere \citep{Finlayson-Pitts2000}, which means that exclusion of $HNO_3$ (formation and removal)/(chemistry) from the chemical mechanism potentially leads to overproduction of ozone in the model, especially at high $NO_2$ concentrations because the competition between $NO_2$ photolysis and its reaction with $OH$ radical ($NO_2 + OH \xrightarrow{M} HNO_3$) is ignored.

@@

\item $NO_3$ radical rapidly photolyzes during the daylight hours, but in nighttime its concentration can increase to measurable levels \citep{Atkinson2008}. This increase in concentration and a lesser availability of the OH radical (which sources are mostly photolytic in nature) make $NO_3$ radical a major contributor to the night chemistry of organics in the troposphere \citep{Finlayson-Pitts2000}. However, exclusion of $NO_3$ radical reactions from our chemical mechanism should not have a huge impact on the results anyway since the mechanism considers mainly decomposition of alkanes with which $NO_3$ radical reacts to a generally minor extent \citep{Atkinson2008}???.

@@

\item Peroxyacetyl nitrate ($CH_3C(O)OONO_2$, PAN) is the most abundant organic nitrate in the troposphere \citep{Hewitt1994}. A classic precursor to the PAN is acetaldehyde ($CH_3CHO$) which oxidation produces PAN through the following reaction sequence:
\item[] $CH_3CHO + OH \rightarrow CH_3CO + H_2O$
\item[] $CH_3CO + O_2 \xrightarrow{M} CH_3C(O)OO$
\item[] $CH_3C(O)OO + NO_2 \xrightarrow{M} CH_3C(O)OONO_2$.
Of all the possible fates for PAN in the atmosphere, thermal decomposition is usually the most important because the rate constant for PAN decomposition strongly depends on temperature. At low temperatures it is quite stable, but at higher temperatures it decomposes giving $NO_2$ and the acetylperoxy radical ($CH_3C(O)OO$) 
\item[] $CH_3C(O)OONO_2 \xleftrightarrow{M} NO_2 + CH_3C(O)OO$.
This temperature dependence has important implications for the role of PAN. When it is formed at lower temperatures (e.g., at high altitude) or transported into colder regions, PAN can act as an $NO_x$ reservoir. Observations of PAN and the sum of reactive nitrogen ($NO_y$) confirm it revealing that PAN can constitute up to 90\% of the total $NO_y$ budget at higher northern latitudes or higher altitudes (e.g., \citep{Jacobi2000}). When an air mass containing PAN is transported into warmer regions, $NO_2$ tied up in PAN can be released back and cause the formation of $O_3$ and $HNO_3$ far away from the original $NO_x$ source. Moreover, if sufficient $NO$ is present, $CH_3C(O)OO$ released from PAN can react in the following manner:
\item[] $CH_3C(O)OO + NO \rightarrow CH_3C(O)O + NO_2$
\item[] $CH_3C(O)O \rightarrow CH_3 + CO_2$
followed by the reaction of $CH_3$ with $O_2$, etc. to form $HCHO$ and $HO_2$, or $HCHO$, $CH_3OH$ and $CH_3OOH$ under low-$NO_x$ conditions. As a result of this chemistry, PAN can act as an accelerator for photochemical smog formation \citep{Finlayson-Pitts2000}. Exclusion of PAN from the chemical mechanism of the model means that we omit the global transport of $NO_x$ provided by PAN, and therefore do not fully estimate tropospheric ozone production. However, since the focus of this research is on the impact of alkyl nitrate chemistry on ozone such a miscount is acceptable due to purely theoretical purpose of the research.
\item Inclusion of all chemistry listed above into the model is a subject for future work.

The constructed chemical mechanism has a limited representation of organic chemistry. It includes (oxidation schemes)/(degradation reactions) of only seven alkanes: methane ($CH_4$), ethane ($C_2H_6$), propane ($C_3H_8$), n-butane ($nC_4H_{10}$), i-butane ($iC_4H_{10}$), n-pentane ($nC_5H_{12}$) and i-pentane ($iC_5H_{12}$). While the oxidation scheme of $CH_4$ is full, oxidation of other six alkanes is not full, but terminated at different oxidation steps.

It has been done due to computational restrictions and  because otherwise the mechanism would 
Another limitation of the model's chemical mechanism is that it does not consider all possible organic reactions (which otherwise could lead to a too large number of reactions but is focused on representing the oxidation of alkanes
full oxidation scheme for methane ($CH_4$) only .

@@

@@
